---
title: "Run Data: 2013-Present"
author: "Joe Martin"
date: "9/12/2021"
output: html_document
---
```{r}
# Briefly describe what the data is, how it was collected, where you will get (or got) it from. How many observations do you have, what was measured? Anything else important to report about the data?
```


The data I intend to use for my project is my personal running data. I started distance running in 2012 just before I got my very first smartphone. When I got my smartphone I started to track all of my runs in the Nike Run Club app. Starting in 2013, this was my primary way of tracking my runs. Toward the end of 2014, I started trying various other apps and mostly got the same data out of them. Apps included MapMyRun and Strava, and possibly a few more. Much of my data from 2015 and 2016 is missing and I’m unsure how to, or if I could ever, recover it. In 2017 I mostly used MapMyRun. In 2018, I moved back to  Nike Running and had a full month in 2020 when I used Strava exclusively. In April of 2020, I bought my first smartwatch (Garmin Forerunner 245), which I’ve been using to track all of my training (running, swimming, and cycling).

I estimate that I have more than 1,000 observations for running. In 2020 when we went into lockdown for COVID-19, I began moving all of my data from my various apps and accounts into a Google sheet. I mainly did this manually. I enter my data into another spreadsheet manually for my most recent data which I primarily use for my training journal. In this journal, I record imprecise numbers – average pace, mile/lap splits, the shoes I wore, weather, time of day/date, total distance, location of my run, distance type (short, medium, or long), and run type (recovery, easy, “steasy,” steady, tempo, track, distance). Along with this, I’ll also take notes about my run, like how I feel, thoughts about how my training is going, whether I’m feeling any pain or note any injuries, etc. With my Garmin watch, I have a Garmin Connect account, which automatically tracks data for me and saves to the cloud. When I export data related to my activities (running, swimming, and cycling all come together), Garmin gives me a .csv file with 36 variables, although some are empty or have 0 for all values. 

Outside of this data, I also have access to data about my sleep and resting heart rate from my Garmin watch. Sleep data includes how long I slept, how long I spent in light sleep, deep sleep, REM sleep, and my respiration during sleep. Resting heart rate data is added below and shows my resting heart rate for each day. I'm interested in using this data and even reviewing activities outside of running (swimming and cycling) to see if they affect my performance, but I don't want to start with such a broad scope. 

I can also export data related to sleep (how long I slept total, length of each sleep stage, respiration), heart rate, steps, stress, and more. I’m interested in exploring this data, but intend to narrow my scope only to my running data for the purpose of this project. 

Going through my data, I have noticed that most of my variables are currently stored as character data. This will be one of the first things I will need to fix to make the most out of my data

```{r}
#At this stage you are not required to already have and show the data, but if you do, even better. Then add a few lines of code which load the data and using some of the commands you learned about, provide summary descriptions of the data.
```

## Historical Data 2013 - 2018
```{r}
# Load Packages
pacman::p_load(pacman, rio, tidyverse, lubridate, here)

# Load Run Data ############################################
# When adding years 2013-2019, change df1 to df20 and make full dataframe out of all years

# 2013 - 2018 Run Data
data13 <- here::here("data", "run2013-18.csv")
df1318 <- read_csv(data13)

# filter out dates without data
df1318 <- df1318 %>% filter(total_time != '')

summary(df1318)

rmarkdown::paged_table(df1318)
```

## Data Combined 2019 - 2021
```{r}
# combined data for 2019-present
data19 <- here::here("data", "run2019.csv")

df19 <- read_csv(data19)
df19$Distance <- as.character(df19$Distance)
df19$`Elev Loss` <- as.character(df19$`Elev Loss`)


data20 <- here::here("data", "run2020.csv")

df20 <- read_csv(data20)
df20$Distance <- as.character(df20$Distance)
df20$`Elev Loss` <- as.character(df20$`Elev Loss`)


data21 <- here::here("data", "run2021.csv")

df21 <- read_csv(data21)
df21$Distance <- as.character(df21$Distance)
df21$`Elev Loss` <- as.character(df21$`Elev Loss`)


df1 <- bind_rows(df19,df20,df21)

summary(df1)

rmarkdown::paged_table(df1)
```

## Resting Heart Rate 4/15/21 - Present
```{r}
rhr <- read_csv("dailyRHR.csv")

rmarkdown::paged_table(rhr)
```

```{r}
# Explain the question you want to answer using the data. What will be your outcome(s) of interest (if any)? What (if any) specific predictors will you focus on? What relations/patterns are you looking for in the data?
```

Outcomes of interest: Pace, heartrate (resting and active), qualitative outcome - positive/negative notes. Variables: shoes, number of track workouts

Through this project, I'm interested in understanding how factors of my training program affect my performance. I would like to see if there are any significant relationships between the frequency of my training and performance, volume (number of miles run per week) and performance, and intensity and performance. I would need to explore my own data more and continue reviewing papers like <a href="https://www.nature.com/articles/s41467-020-18737-6">Human running performance from real-world big data" by Thorsten Emig & Jussi Peltonen</a>. I would think specific predictors would include average pace, average/max heartrate during workouts, and resting heartrate.

```{r}
#As much as you know, suggest how you will analyze it. At this stage in the course, we haven’t covered analysis approaches yet, so you can keep things vague and non-technical here. 
```
In order to search for relationships and patterns in my data, I will first need to identify specific events - like races or long runs - and compare weeks of training against each other. I do one long distance run just about every week. My definition of "long distance" has changed as I began training more and improved. 

The other way I may go about analyzing patterns in my data would be by picking specific distances described the same way (easy, steady, tempo). For example, a common workout for me would be an up-tempo 10k (6.22 miles) or an easy 8-miler. 

```{r}
# You are allowed, but not yet required, to provide background information for the question you plan to answer. For instance you can describe why it’s an interesting question, who else has done similar analyses, how your analysis will be new/different, etc. Similar to what you read in an introduction to a research paper. For the final report, you’ll need these parts. For part 1, they are not required, but you are welcome to already scribble down some of that.
```

Coaches are expensive to hire. Apps provide users with some basic stats about their runs, but users typically have to purchase a subscription to get useful insights into training. These insights aren't catered to individual users. Programming languages give individuals an opportuinity to gain deep insights into their own training. "Human running performance from real-world big data" by Thorsten Emig & Jussi Peltonen is a good paper. There is a lot of research about performance, but much research is conducted for individual athletes, and is therefore not accessible outside of organizations.



```{r}
# Eventually, for your final report, what you write for this part will go into different sections of the full report. Some will go into the introduction, some in the methods section. You can already place these items there, or for now just write them as a single section.
```